---
title: "Untitled"
format: html
---

# Introduction

This report was last generated on `{r} Sys.Date()`.

## About the data
Navigate to this dataset BGchem2008data.csv by Craig Tweedie that is published on the Arctic Data Center. Craig Tweedie. 2009. North Pole Environmental Observatory Bottle Chemistry. Arctic Data Center. doi:10.18739/A25T3FZ8X.

# Setup

```{r load packages}
#| warning: false
#| message: false
library(here)
library(readr)

```

```{r}
#| message: false
bg_chem <- read_csv(here::here("data/BGchem2008data.csv"))

```

# Analysis

## Summary Statistics

```{r}
#| eval: false
#| echo: false
colnames(bg_chem)
str(bg_chem)
head(bg_chem)
summary(bg_chem)
```

```{r}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm     <- mean(bg_chem$NH4)
phos    <- mean(bg_chem$P)
```

## Redfield Ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
```

The Redfield ratio for this dataset is approximately `{r} round(ratio)`.

## Plot

```{r}
plot(bg_chem$P, bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4)
```

# Conclusion

This is the end.
