---
title: "Working with Vector Spatial Data in R"
author: "Ann Holmes"
format: html
---

## Setup
```{r}
# Data wrangling packages:
library(readr)
library(dplyr)
library(tidyr)
library(here)

# Simple Features vector spatial data:
library(sf)

# Data visualization packages:
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
```

### Download data
```{r}
out_dir <- here::here('data/ak_regions')

if(!dir.exists(out_dir)) {
    knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

    download.file(url = knb_url, 
                  destfile = 'shapefile_demo_data.zip', 
                  mode = 'wb')

    unzip('shapefile_demo_data.zip', exdir = out_dir)

    file.remove('shapefile_demo_data.zip') #removes the zipfile
}
```

##### add: ignore alaska spatial files to gitignore data/ak_regions/

#### Population data
```{r}
out_file <- here::here('data/alaska_population.csv')

if(!file.exists(out_file)) {
    pop_data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3A7fc6f6db-c5ea-426a-a743-1f2edafb43b8"

    download.file(url = pop_data_url, 
                  destfile = out_file, 
                  mode = 'w')
}
```

## Data Exploration

### Read in data
#### Spatial data
```{r}
ak_rgns_sf <- read_sf(here("data/ak_regions/ak_regions_simp.shp"))

# plot(ak_rgns_3338_sf) ### splits AK across left and right margins
```

#### Population data
```{r}
pop_df <- read_csv(here("data/alaska_population.csv")) %>%
    filter(Year == 2015) %>%
    select(city, lat, lng, population = total) %>%
    drop_na()
```

### Transform data 

#### Transfor spatial data for plotting
```{r}
ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

# st_crs(ak_rgns_3338_sf) ### run in console to check

plot(ak_rgns_3338_sf) #now AK is centered
```

#### Transform pop_df into a spatial object
```{r}
pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'), #lat and long were previously considered as numbers
                        crs = 4326,
                        remove = FALSE)
```

#### Join population data with AK regions
```{r}
pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)

pop_joined_sf <- st_join(pop_3338_sf, 
                         ak_rgns_3338_sf, 
                         join = st_within)

head(pop_joined_sf)
```

### Explore object class
```{r}
class(ak_rgns_3338_sf)
head(ak_rgns_3338_sf)
glimpse(ak_rgns_3338_sf)
```

## Analysis

### Calculate total population by region
```{r}
pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>% #removes the spatial data
    group_by(region) %>%
    summarize(total_pop = sum(population))
```

#### Connect (non-spatial) population df back to the AK regions spatial df
```{r}
pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf, 
                             pop_rgn_df, 
                             by = "region") #join by region name not spatial data

### plot to check
plot(pop_rgn_3338_sf["total_pop"])
```

## Visualization
