---
title: "Data wrangling"
format: html
engine: knitr
---
# Clean and wrangle data

## Setup

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## Source data

### Read data

```{r}
catch_original <- readr::read_csv(file = "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

```{r}
colnames(catch_original)
head(catch_original)
summary(catch_original)
```

## Clean and reshape data

```{r}
catch_data <- catch_original %>%
    dplyr::select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)
dplyr::glimpse(catch_data)
```

### Change column from character to numeric

```{r}
# Make the Chinook column numeric
catch_clean <- catch_data %>%
    dplyr::mutate(Chinook = as.numeric(Chinook))
```

### Check for NAs
```{r}
(i <- which(is.na(catch_clean$Chinook)))
catch_data[i, ]
```

### Use a conditional to fix the bad value, then make the column numeric and check result
```{r}
catch_clean <- catch_data %>%
    dplyr::mutate(
        Chinook = dplyr::if_else(condition = Chinook == "I", 
            true = "1",
            false = Chinook),
        Chinook = as.numeric(Chinook))

catch_clean[i, ]
```

### Change shape of data to suit our needs
#### Pivot all columns except Region and Year into long format, and check result

```{r}
catch_long <- catch_clean %>% 
    tidyr::pivot_longer(cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch_thousands")
head(catch_long)
```

#### Pivot the data back into wide format
```{r}
catch_wide <- catch_long %>%
    tidyr::pivot_wider(names_from = species,
        values_from = catch_thousands,
        values_fill = 0)
head(catch_wide)
```

### Convert to fish numbers

#### Convert to actual fish numbers, then remove the superseded column of catch in thousands

##### Advice: don't reuse an object name like this ⚠️⚠️⚠️ (could use catch_long_v2)

```{r}
catch_long <- catch_long %>%
    dplyr::mutate(catch = catch_thousands * 1000) %>% 
    dplyr::select(-catch_thousands)
dplyr::glimpse(catch_long)
head(catch_long)
```
## Summarize data

### Calculate summary statistics

#### Group by region, then calculate averages within the established groups
```{r}
mean_region <- catch_long %>%
    dplyr::group_by(Region) %>%
    dplyr::summarize(mean_catch = mean(catch, na.rm = TRUE), #removes NA before taking mean
        .groups = "drop")

dplyr::glimpse(mean_region)
head(mean_region)
```

#### Limit catch to the SSE region and check result

```{r}
sse_catch <- catch_long %>%
    dplyr::filter(Region == "COP")

unique(sse_catch$Region)

head(sse_catch)
```

#### Now arrange mean_catch column by descending value

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch, na.rm = TRUE), #removes NA before taking mean
        .groups = "drop")%>%
    arrange(desc(mean_catch))

dplyr::glimpse(mean_region)
head(mean_region)
```

### As an example, split single columnsingle column into numerous columns (not catch data)

```{r}
sites_df <- data.frame(site = c("HAW-101", "HAW-103", "OAH-320", "OAH-219", "MAU-039"))

head(sites_df)
```

#### Split site by the dash between the site letters and its code

```{r}
#| label: split data
sites_df %>%
    tidyr::separate_wider_delim(cols = site, delim = "-",
        names = c("island", "site_number"), cols_remove = FALSE)
```
